---
title: "Protein Coverage of Different Proteolysis Enzymes"
runtime: shiny
output: html_document
---
```{r, eval=FALSE, echo=FALSE}
# To run this file, open in RStudio and click "Run Document".  It should open in a browser, or RStudio preview Browser and allow you to
# select files of peptide lists and view the results as venn diagrams.
```

```{r,message=FALSE, echo=FALSE}
library (data.table)
library (VennDiagram)
library (grid)
library (RColorBrewer)
library (R.utils) # required for reading gzip files in fread

LoadPeptideSites = function(file){
  peptides = fread(file)[,.(protein = `Leading razor protein`, start = `Start position`, end = `End position`)]
  peptides = peptides[!substr(protein,1,3) %in% c("CON", "REV") ,]
  if (any(!complete.cases(peptides))){
    showNotification(sprintf("Missing values detected in %d rows", sum(!complete.cases(peptides))), type="warning")
    #warning (sprintf("Missing values detected in %d rows", sum(!complete.cases(peptides)), basename(file)))
    peptides <- peptides[complete.cases(peptides)]
  }

  
  sites = do.call(rbind.data.frame, 
                  lapply(1:nrow(peptides), 
                         function(i) {
                           data.frame(protein = peptides$protein[i], 
                                      site=peptides$start[i]:peptides$end[i], 
                                      stringsAsFactors=FALSE)
                         }
                  )
  )
  return (unique(paste (sites$protein, sites$site, sep="-") ))
}


```


```{r, echo = FALSE}
files = c()

```

Select files of peptide lists to "upload" and process. 

Expectations of files:

1. All peptides from a single enzyme (or other proteolysis condition to compare) in a single file.
2. File name will include an enzyme name (or other descriptor you'd like to use in images) as a prefix followed by an underscore, e.g. Trypsin_Peptides.txt
2. File contents are a text table (gzip ok), with at least the columns:  "Leading razor protein", "Start position", "End position".

All other columns will be ignored.  Protein names that begin with CON or REV will be ignored.



```{r, echo = FALSE}
#Input elements go here

#htmlOutput("enzyme.list.out")

fileInput("file1.input", "select your enzyme files", multiple=TRUE)
```
A table of loaded enzymes will appear below
(this takes a few seconds to load after selecting files):
```{r, echo=FALSE}
renderTable ({
  req(values$siteSets)
  data.frame ("Enzyme" = names(values$siteSets), "Count_Sites" = unlist(lapply(values$siteSets, length)))
})

```

```{r, echo = FALSE}
siteSets = list()
values = reactiveValues(siteSets = siteSets)

observeEvent(input$file1.input, {
  withProgress(message="Building non-redundant sets of sites", value = 0.1,{
    for (i in seq_along(input$file1.input[,1])){
      enzymeName = unlist(strsplit(basename(input$file1.input$name[i]), split="_"))[1]
      tryCatch({
          values$siteSets[[enzymeName]] <- LoadPeptideSites(input$file1.input$datapath[i])
        }, 
        error = function(e){
          showNotification(paste ("Error while reading:", basename(input$file1.input$name[i]), e, sep="\n"), type="error")
          return
        })
      incProgress(amount = 1/nrow(input$file1.input))
    }
  })
})

reactivePalette = reactive({
    palette = brewer.pal(8, "Set2")
    names(palette)[seq_along(values$siteSets)] = names(values$siteSets)
    palette
})

```

## Single Venn Diagrams
Numbers are the number of unique amino acid positions included in the entire peptide table.
```{r, echo=FALSE}

singleVImageHeight = function(){200}
singleVImageWidth = function(){
  imageCount <- length(names(values$siteSets))
  if (imageCount == 0) imageCount <- 1
  singleVImageHeight() * imageCount
}

renderPlot(height = singleVImageHeight, width = singleVImageWidth,{
  req(values$siteSets)
  req (length(values$siteSets) >=1)

  palette = brewer.pal(8, "Set2")
  names(palette)[seq_along(values$siteSets)] = names(values$siteSets)
  
  singleVenns = sapply (names(values$siteSets), 
                        function(name){
                          draw.single.venn(area=length(values$siteSets[[name]]), category = name, ind=FALSE, fill = reactivePalette()[name])
                        },
                        simplify=FALSE
  )
  
  maxSet = max(sapply (values$siteSets, length))
  
  for (i in 1:length(values$siteSets)){
    scaleFactor = sqrt(length(values$siteSets[[i]])/maxSet)
    pushViewport ( viewport(x=(i-1)/length(values$siteSets), y=0, height=scaleFactor, width=scaleFactor * 1/length(values$siteSets), just=c("left", "bottom")))
    grid.draw(singleVenns[[i]])
    popViewport()
  }
  
})

```

## Double Venn Diagrams

```{r, echo=FALSE}

doubleVImageHeight = function(){250}
doubleVImageWidth = function(){
  imageCount <- if (length(values$siteSets) < 2)
    0
  else
    length(combn(names(values$siteSets), 2, simplify=FALSE))
  if (imageCount == 0) imageCount <- 1
  doubleVImageHeight() * imageCount
}


renderPlot(height = doubleVImageHeight, width=doubleVImageWidth,{
  req(values$siteSets)
  req (length(values$siteSets) >=2)
  pairs = combn(names(values$siteSets), 2, simplify=FALSE)
  maxSet = max (sapply (pairs, 
                        FUN = function(pair){
                          length(union(values$siteSets[[pair[1]]], values$siteSets[[pair[2]]]))
                        }
  )
  )
  grid.newpage()
  i = 0
  for (pair in pairs){
    i = i+1
    set1 = values$siteSets[[pair[1]]]
    set2 = values$siteSets[[pair[2]]]
    area1 = length(set1)
    area2 = length(set2)
    cross.area = length(intersect(set1,set2))
    scaleFactor = sqrt(length(union(set1,set2))/maxSet)
    pushViewport ( viewport(x=(i-1)/length(pairs), y=0, height=scaleFactor, width=scaleFactor * 1/length(pairs), just=c("left", "bottom")))
    
    draw.pairwise.venn (area1, area2, cross.area, category = pair, fill = reactivePalette()[pair])  
    
    popViewport()
  }
  
  
})
```

## Triple Venn Diagrams

```{r, echo = FALSE}
tripleVImageHeight = function(){300}
tripleVImageWidth = function(){
  imageCount <- if (length(values$siteSets) < 3)
    0
  else
    length(combn(names(values$siteSets), 3, simplify=FALSE))
  if (imageCount == 0) imageCount <- 1
  tripleVImageHeight() * imageCount
}
renderPlot(height = tripleVImageHeight, width = tripleVImageWidth, {
  req(values$siteSets)
  req (length(values$siteSets) >=3)
  triples = combn(names(values$siteSets), 3, simplify=FALSE)
  maxSet = max (sapply (triples, 
                        FUN = function(triple){
                          length(union(union(values$siteSets[[triple[1]]], 
                                             values$siteSets[[triple[2]]]),
                                       values$siteSets[[triple[3]]]
                          )
                          )
                        }
  )
  )
  grid.newpage()
  i = 0
  for (triple in triples){
    i = i+1
    set1 = values$siteSets[[triple[1]]]
    set2 = values$siteSets[[triple[2]]]
    set3 = values$siteSets[[triple[3]]]
    area1 = length(set1)
    area2 = length(set2)
    area3 = length(set3)
    n12 = length(intersect(set1,set2))
    n23 = length(intersect(set2,set3))
    n13 = length(intersect(set1, set3))
    n123 = length(intersect(intersect(set1,set2), set3))
    scaleFactor = sqrt(length(union(union(set1,set2),set3))/maxSet)
    pushViewport ( viewport(x=(i-1)/length(triples), y=0, height=scaleFactor, width=scaleFactor * 1/length(triples), just=c("left", "bottom")))
    
    draw.triple.venn (area1, area2, area3, n12, n23, n13, n123, category = triple,  fill =reactivePalette()[triple] ) 
    
    popViewport()
  }
})

```
