---
title: "R Notebook"
output: html_notebook
---
This template file located at:
/Applications/RStudio.app/Contents/Resources/resources/templates/notebook_osx.Rmd

```{r}
library (data.table)
library (ComplexHeatmap)
library (ggplot2)
rotate.x.axis.text <- theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))

source("../../bp_utils/ManageScriptData.R")
WriteSessionInfo()
WriteInstalledPackages()
```

# path to results.txt file, MSstats GroupComparison output
```{r}
results.txt <- "/Users/ben/Box/Project_Coronavirus/00_data/proteomics/Level2/UK4_PH_results_v2.txt"
species <- "HUMAN"  # or "MOUSE"

githubKroganLab <- "../../"
source(file.path(githubKroganLab,  "bp_utils" ,"KinaseActivityScores.R"))



filterResults <- function(x){
  return(x)
}


filterResults <- function(results){
  return (results[grepl( "(VIC|IC19)_(.+)-Mock_\\2", Label), ])
}  

```



# Volcanoes


```{r, fig.width=5, fig.height=4}
results <- filterResults (fread (results.txt))

pvalueThreshold <- 0.05
log2FCThreshold <- 1
pvalueVariable <-  "adj.pvalue" # or "pvalue"


results[,sig := "not"]
results[results[[pvalueVariable]] < pvalueThreshold & abs(log2FC) > log2FCThreshold, sig := ifelse(log2FC  <0, "down", "up")]
results[, yVariable := -log10(results[[pvalueVariable]])]

p <- ggplot (results, aes(x=log2FC, y = yVariable, color = sig)) + 
  geom_point(show.legend = FALSE, alpha = 0.5, size = 1, shape = 16) + 
  facet_wrap (~Label) + 
  scale_color_manual(values = c(down= "blue", not = "gray", up = "red")) +
  #scale_x_continuous(limits = c(-4,4)) +
  scale_y_continuous(name = paste(pvalueVariable, "(-log10)", collapse = "")) +
  geom_hline(yintercept = -log10(pvalueThreshold), lty="dotted") +
  geom_vline(xintercept = c(-1, 1) * log2FCThreshold, lty="dotted") + 
  theme_bw() 


BackupAsPDF (p, "Volcanos_")


results[,c("posGroup", "negGroup") := tstrsplit(Label, split = "-")]

p <- ggplot (results, aes(x=log2FC, y = yVariable, color = sig)) + 
  geom_point(show.legend = FALSE, alpha = 0.5, size = 1, shape = 16) + 
  facet_grid(rows=vars(posGroup), cols = vars(negGroup)) + 
  scale_color_manual(values = c(down= "blue", not = "gray", up = "red")) +
  #scale_x_continuous(limits = c(-4,4)) +
  scale_y_continuous(name = paste(pvalueVariable, "(-log10)", collapse = "")) +
  geom_hline(yintercept = -log10(pvalueThreshold), lty="dotted") +
  geom_vline(xintercept = c(-1, 1) * log2FCThreshold, lty="dotted") + 
  theme_bw() 


BackupAsPDF (p, "Volcanos_grid_")


```

## barcharts

```{r}
#results[, sig := factor(sig, levels= c("up", "not", "down"))]
summaryData <- results[, .(count = length(unique(Protein))), by = .(Label, sig) ]
summaryData[, sig := factor(sig, levels= c("up", "not", "down"))]
setorder(summaryData, Label, -sig)
summaryData[, textY := cumsum(count), by = Label]
summaryData[, textY := textY - count/2] # move down to center of bar part
p <- ggplot(summaryData, aes (x = Label, fill = sig, y = count, label = count))  +
  scale_fill_manual(values = c(down= "blue", not = "gray", up = "red")) +
  geom_col() +
  geom_label(aes(y = textY), color = "white")  +
  theme_bw()

BackupAsPDF(p, "BarChart_CountSigPeptideGroups_")
```


# Kinase analysis

```{r}
kinaseData <- loadKinaseDataOmniPath(species = species)


#load MSstats results data
#results.txt <- file.path(paste0(outFilePrefix, "_output", collapse=""), "results.txt")
results <- filterResults(fread (results.txt)) 

# convert to single site info based on gene names
source(file.path(githubKroganLab,  "bp_utils" ,"UniprotIDMapping.R"))
results[,gene := multiUniprotSites2multiGeneSites(Protein)]
singleSiteResults <- prepare_AMSS_ResultsFile(results, column = "gene")

labels <- unique(singleSiteResults$Label)
#kinAct <- kinaseActivity (expanded2SingleSites[Label == labels[4]])

kinActList <- lapply (labels, FUN=function(lab){kinaseActivity(singleSiteResults[Label == lab & representative==TRUE],
                                                               plots = FALSE,
                                                               kinaseData = kinaseData)})
names(kinActList) <- labels

kinActFull.scores <- rbindlist(lapply(kinActList, FUN = function(x)x$scores), idcol="Label")
kinActFull.mapped <- rbindlist(lapply(kinActList, FUN = function(x)x$kinaseMapped)) # Label is already in these tables

kinaseSummaryScores.csv <- ScriptAndDatedFileName("kinaseSummaryScores.csv")
kinaseSubstrateData.csv <- ScriptAndDatedFileName("kinaseSubstrateData.csv")
message (sprintf("Writing kinase output files to\n\t%s\n\t%s", kinaseSummaryScores.csv, kinaseSubstrateData.csv))

fwrite (kinActFull.scores, kinaseSummaryScores.csv)
fwrite (kinActFull.mapped, kinaseSubstrateData.csv)

```

## Heatmap of significant kinases
```{r, fig.width = 4, fig.height= 5}
sigKinases <-  kinActFull.scores[fdr.BH < 0.05 & N >= 2, unique(CTRL_GENE_NAME)]

sigKinase.mat.z <- as.matrix(dcast (kinActFull.scores[CTRL_GENE_NAME %in% sigKinases], CTRL_GENE_NAME~Label, value.var = "Z"),
                              rownames = "CTRL_GENE_NAME")

sigKinase.mat.N <- as.matrix(dcast (kinActFull.scores[CTRL_GENE_NAME %in% sigKinases], CTRL_GENE_NAME~Label, value.var = "N"),
                                  rownames = "CTRL_GENE_NAME")


hm <- Heatmap (sigKinase.mat.z, 
         cluster_columns=TRUE, 
         name = "Kinase Z Score",
         column_split = tstrsplit(colnames(sigKinase.mat.z), split="[-_.]")[[1]],
         col = circlize::colorRamp2(breaks = c(-3, 0, 0, 3), colors = c("blue", "gray", "gray", "red")),  # two midpoints in case you want to have a wider gray bar around 0
         cell_fun = function(j, i, x, y, width, height, fill) {
                                  if (!is.na(sigKinase.mat.N[i,j])){
                                    grid.text(sprintf("%.0f", sigKinase.mat.N[i, j]), x, y, gp = gpar(fontsize=10, col="white"))
                                  }
                                })

BackupAsPDF(hm, prefix = "Heatmap_SigKinases")

hm <- Heatmap (sigKinase.mat.z, 
         cluster_columns=FALSE, 
         name = "Kinase Z Score",
         column_split = tstrsplit(colnames(sigKinase.mat.z), split="[-_.]")[[1]],
         col = circlize::colorRamp2(breaks = c(-3, 0, 0, 3), colors = c("blue", "gray", "gray", "red")),  # two midpoints in case you want to have a wider gray bar around 0
         cell_fun = function(j, i, x, y, width, height, fill) {
                                  if (!is.na(sigKinase.mat.N[i,j])){
                                    grid.text(sprintf("%.0f", sigKinase.mat.N[i, j]), x, y, gp = gpar(fontsize=10, col="white"))
                                  }
                                })

BackupAsPDF(hm, prefix = "Heatmap_SigKinases_noColCluster")


```


## plots of individual kinase substrates


```{r, fig.width=6 fig.height=6}
p <- BarplotKinaseActivities(kinActFull.scores, kinActFull.mapped, sigKinases = sigKinases)
BackupAsPDF(p, "SigKinaseSites_BoxPlots_ByLabel_")

p <- BarplotKinaseActivities(kinActFull.scores, kinActFull.mapped, sigKinases = sigKinases, reverse=TRUE)
BackupAsPDF(p, "SigKinaseSites_BoxPlots_ByKinase_")

```

# Enrichment analysis
## Prepare sets of proteins from our data
```{R}
results <- filterResults(fread (results.txt))

pvalueThreshold <- 0.05
log2FCThreshold <- 1
pvalueVariable <-  "adj.pvalue" # or "pvalue"


results[,sig := "not"]
results[results[[pvalueVariable]] < pvalueThreshold & abs(log2FC) > log2FCThreshold, sig := ifelse(log2FC  <0, "down", "up")]

# multiproteins will be excluded from enrichment by the enrichment process itself 
# because (e.g.) the string "A0A0B4J2F2;P57059" does not match any uniprot exactly (likewise for genes)
results[, uniprot := strsplit(gsub( "_[STY][0-9]+(;|$)", "\\1",  Protein),";") |>
          lapply(FUN = unique) |>
          sapply(FUN = paste, collapse = ";")]


universe <- results$uniprot |> unique()

```


## load gene ontology data
```{R}
source (file.path(githubKroganLab, "bp_utils", "enrichmentTestFunctions.R"))

gmt <- loadGmtFromBioconductor(dbName = "org.Hs.eg.db", ontology = "ALL", keyType = "UNIPROT")

```

```{r}
enpl.out <- enrichmentOnGroupsPL(groupTable = results[sig!="not"], geneColumn = "uniprot", groupColumns = c("Label", "sig"), gmt = gmt, universe = universe)
```




## plot
```{r, fig.height=4, fig.width=5}

a <- enrichHeatmapBestPerGroup(enpl.out$simp[[1]], enpl.out$simp[[2]],
                          #cols = sort(c(negCols, posCols)),
                          negCols = grep ("down", unique(enpl.out$simp[[1]]$Label.sig), value = TRUE),
                          groupColumn = "Label.sig",
                          topN = 8,
                          cluster_columns=FALSE, 
                          reduceRedundantsAcrossGroups=TRUE,
                          minCount = 1,
                          max_pAdjust = 0.05)

# to draw results to a file; you will want to play with height and width to get reasonably shaped plot
# cairo_pdf("My_enrichment_heatmap_file.pdf", height=6, width=8)  # or replace cairo_pdf with pdf if you don't have cairo installed
# draw (a$hmList)
# dev.off()


BackupAsPDF(a, "PH_Protein_enrichments")
```


