---
title: "R Notebook"
output: html_notebook
---
```{r}
library (data.table)
library (ggplot2)
```

# This has been replaced by 2025_04_17_NetworkPropagation.Rmd

```{r}
stop( "This has been replaced by file 2025_04_17_NetworkPropagation.Rmd. Proceed at your own risk (as always)")
```


# Load data from MSstats
I will use log2FC (magnitude) and pvalue (significance) combined to generate a heat (magSigScore).  

```{r}

results.txt <- "../../bp_utils/data/short_ph_results.csv.gz"
results <- fread (results.txt)


magSigScore <- function(log2FC, pvalue, magnitudeScale = 2){
  magnitude <- magnitudeScale * abs(log2FC)
  significance <- -log10(pvalue)
  significance <- ifelse (significance > magnitude, magnitude, significance)
  sqrt(magnitude * significance)
}


results[, magSig := magSigScore(log2FC, pvalue)]
# remove outliers and infinites
results <- results[abs(log2FC) < 10,]

```


# deal with phospho data
Above data source is phospho data, so multiple values exist per protein (multiple sites per protein).  Network propagation needs only 1 score per protein.
This hack removes the site info per protein, discards the multiproteins, and then takes the max score per protein

```{r}

results[, uniprot := gsub("_[STY][0-9]+", "", Protein) %>% 
          strsplit (";") %>% 
          lapply (unique) %>% 
          lapply (paste, collapse = ";") %>%
          unlist]


results <- results[!grepl(";", uniprot)]

proteinHeatPerLabel <- results[!is.na(magSig), .(heat = max(magSig, na.rm = TRUE)), by = .(Label, uniprot)]

source ("../../bp_utils/UniprotIDMapping.R")
# I use genes here as my identifiers.  Not necessarily best...
proteinHeatPerLabel[, gene := translateUniprot2GeneName(uniprot, species = "HUMAN")] # or "MOUSE" or "RAT"

```

# Load omnipath network and build network-specific S_matrix 
```{R}
interactions <- OmnipathR::import_omnipath_interactions(organism = 9606) #9606 human, 10090 mouse; 10116 rat

setDT(interactions)


```

## use igraph to get largest connected component.
```{r}
# If I was using uniprots, I'd select `source` and `target` instead of source_genesymbol etc
# these will end up as row/column names on the S_matrix
edges <- interactions[, .(source_genesymbol, target_genesymbol)]

net.graph <- igraph::graph_from_edgelist(as.matrix(edges ), directed= FALSE)
components <- igraph::decompose(net.graph)
componentSizes <- sapply (components, FUN = function(x)length(igraph::V(x)))
barplot(componentSizes)
net.graph <- components[[which.max(componentSizes)]]
length(igraph::V(net.graph))
length(igraph::E(net.graph))

```

## output to prepare for python

```{r}
fwrite (setDT(igraph::as_data_frame(net.graph))[, .(from ,to)], "OmniPath_Interactions_maxComponent.csv")

```

## python code to calculate S matrix
This is much faster in python, and you only need to do it once per network (don't need to repeat when your MS data changes)
This calculates a different S matrix fo different values of `pr` (the restart probability).  See the for loop at line 145 to change these
possible values of pr.

```{python3, eval = FALSE}
import networkx as nx
import numpy as np
import pandas as pd
import time

# For math guidance/explanation
# see https://doi.org/10.1038/nrg.2017.38
#  Network propagation: a universal amplifier of genetic associations
# AU  - Cowen, Lenore
# AU  - Ideker, Trey
# AU  - Raphael, Benjamin J.
# AU  - Sharan, Roded
# PY  - 2017
# DA  - 2017/09/01
# TI  - Network propagation: a universal amplifier of genetic associations
# JO  - Nature Reviews Genetics
# 
# This uses an S matrix (terminology from above paper) which is a scaled transpose of 'inv_denom' used elsewhere in Krogan lab (see Mehdi's code)
# to propagate with an S matrix do: S %*% heat_0
# to propagate with inverse_denom do: heat_0 %*% inv_denom * pr
# Note:
# ! Order around %*% matters and S includes the pr factor !
# col sums of S = 1
# row sums of inv_denom = 1/pr 


edges = pd.read_csv ("OmniPath_Interactions_maxComponent.csv", sep = ",")
filteredEdges = edges[['from','to']]#mouseEdges[mouseEdges['combined_score'] > 600][['protein1','protein2']]
edge_list = filteredEdges.values.tolist()

# Create Graph
G = nx.Graph(edge_list)
G.remove_edges_from(nx.selfloop_edges(G)) # removing selfloops
largest_cc = max(nx.connected_components(G), key=len) # only taking the largest connected component
Gc = G.subgraph(largest_cc)

# Calculate matrix inputs
nodes = Gc.nodes()
A = nx.adjacency_matrix(Gc, nodelist=nodes)
D = np.array([x[1] for x in nx.degree(Gc)]) # degree of each node

# integrity check: node order matches
degreeNodes = np.array([x[0] for x in nx.degree(Gc)])
assert (all(degreeNodes == nodes))

Dinverse = np.diag(1/D)
W = np.matmul(A.todense(), Dinverse)
I = np.identity(W.shape[0])

for pr in (0.2, 0.3, 0.5):
  print ("starting matrix inverse calculation at {}".format(time.asctime(time.localtime())))
  S_matrix = pr * np.linalg.inv(I-(1-pr)*W)
  print ("finished matrix inverse calculation at {}".format(time.asctime(time.localtime())))
  np.save("S_matrix.omnipathInteractions.pr%0.2f.npy"%(pr), S_matrix)
  # node names
  nodes_network = pd.DataFrame(nodes) # Getting node names(gene names)
  nodes_network.rename(columns={0: "gene_name"}, inplace=True) # rename column header
  nodes_network.to_csv("S_matrix.omnipathInteractions.pr%0.2f.nodeNames.csv" %(pr), index=False)
  
  
```



# network prop

```{r}
source ("../../bp_utils/NetworkPropagation.R")
```


## Load S Matrix
```{r, fig.width = 10, fig.height = 8}
library (RcppCNPy)
S_matrix <- LoadNumPyS_matrix("S_matrix.omnipathInteractions.pr0.20.npy", 
                              "S_matrix.omnipathInteractions.pr0.20.nodeNames.csv") 

# take a gander at a portion of hte S_matrix
#ComplexHeatmap::Heatmap(log10(S_matrix[1:1000, 1:1000]), show_row_names = FALSE, show_column_names = FALSE) %>% BackupAsPDF()
```


## network propagation on each "Label" independently
```{r}
# NetworkPropagateS_matrix requries a gene and heat column in the data.table passed as second argument
# if you are using uniprots instead of gene symbols, just rename the column to 'gene' 


groups <- unique(proteinHeatPerLabel$Label)
names(groups) <- groups
allNP <- pbapply::pblapply (groups, function(g) NetworkPropagateS_matrix(S_matrix,
                                                                         proteinHeatPerLabel[Label == g], # requires gene and heat columns
                                                                         networkHeatOnly = TRUE,
                                                                         permuteOnlyInObserved=TRUE,
                                                                         numPermutations = 20000))
```

## data table from list output

```{r}

allNPdt <- rbindlist(allNP, idcol = "Label")
fwrite (allNPdt,  "AllNetworkPropagation_OmniPath_Interactions.rs0.2.csv")


p <- ggplot (allNPdt, aes(x = pvalue)) + geom_histogram(boundary = 0.0, binwidth = 0.005) + facet_wrap(~Label)
p

```


